Does freshwater connectivity influence phosphorus retention in lakes?

### 2019-01-23 manuscript/manuscript.Rmd
add table captions

\begin{figure}
  \begin{center}
    \includegraphics[width=0.45\textwidth, height=3.5in,keepaspectratio]{../figures/01_conceptual_p-cycle_pt2}
  \end{center}
  \caption{Major lake and watershed factors affecting lake P retention. Shaded symbols indicate factors typically considered in P retention models whereas open symbols indicate additional factors specific to the present study. Dashed lines indicate inferred relationships, which cannot be tested with available data, but are otherwise discussed herein.}\label{fig1}
\end{figure}

@@

  \begin{center}
    \includegraphics[width=0.5\textwidth,keepaspectratio]{../tables/crop/01_lake_characteristics_table} \label{table1}
  \end{center}

@@

In addition to data from the NES and LAGOS-NE datasets, we calculated several connectivity metrics that we expected would be related to lake P retention (Figure \ref{fig2}). Some of these metrics were stream-based with the goal of capturing aspects of the configuration of each lakes’ upstream surface water network (Figure \ref{fig2}C-D). In particular, we chose metrics that would quantitatively approximate network complexity under the assumption that highly complex networks are also low connectivity networks. This assumption is supported by the findings of stream network simulations where increased network complexity leads to increased network resistance and ultimately decreases in network connectivity \citep{rodriguez2001fractal}. In addition to stream-based metrics, we calculated lake-stream-based metrics that we expected would reflect the likelihood of P trapping in upstream lakes prior to arriving at a given focal lake via tributary flow (Figure \ref{fig2}A, 2B). A simple metric that captures this likelihood is the presence (or absence) of an upstream lake (greater than 4 ha) which we define as “lake connection” \citep{fergusFreshwaterLandscapeLake2017}. In addition to lake connection, we calculated related metrics such as total upstream lake area and the network distance to the closest upstream lake.
To examine the importance of spatial extent relative to our connectivity metrics, we calculated each metric at multiple extents (Figure \ref{fig3}). First, we calculated connectivity metrics at the scale of individual lake subwatersheds. We defined a lake subwatershed as the area draining into a particular lake exclusive of any upstream areas that drain into a lake greater than or equal to 10 ha (0.1 $km^{2}$). Next, we calculated connectivity metrics at the scale of entire upstream lake networks (lake watershed extent). We defined a lake watershed as the area draining into any part of the upstream network irrespective of the presence or absence of upstream lakes (Figure \ref{fig3}).
All connectivity metrics were calculated using the high-resolution National Hydrography Dataset (NHD) as a primary input \citep{usgsNationalHydrographyDataset2018}. Average link length was calculated as the total stream length in a given watershed divided by the number of stream reaches after dissolving (removing) any network points that do not occur at a stream junction. Stream density was calculated as the length of all streams in the watershed (minus artificial lines through lakes) expressed in units of meters per hectare. Upstream lake area was calculated as the sum of the lake area in the upstream watershed expressed in square meters. Stream order ratio was defined as the number of headwater (first-order) streams in the upstream watershed of the focal lake divided by the total number of higher order (> 1) streams \citep{labarberaFractalDimensionStream1989}. Closest distance to an upstream lake was defined as the shortest path-distance (rather than the straight-line distance) to a lake upstream from the focal lake. We calculated all connectivity metrics and lake watershed extents using the `streamnet` and `nhdR` R packages respectively \citep{R-streamnet, R-nhdR}. The algorithms in the `streamnet` package use the `sf` R package \citep{pebesma2018} as well as the `v.net` and `v.stream.order` modules \citep{jasiewiczNewGRASSGIS2011} included in GRASS GIS \citep{GRASS_GIS_software}. All processed connectivity data and code are available at [Zenodo DOI].

@@

We modelled lake total phosphorus retention (hereafter, P retention) using the Vollenweider equation that models P retention as a function of water residence time and a parameter (k) conceptually representing in-lake P decay \citep{chapra2008surface, vollenweiderInputoutputModels1975}. Although there are several variants of this basic equation, we selected a 2-parameter form (equation 1) that has been
shown to have good performance in multiple cross-sectional studies \citep{brett2008, cheng2010}:

$$
Ri = 1 - \frac{1}{1 + kt^{x}_{i}}
$$

where $R_i$ is P retention as a fraction of P inputs, i $\tau$ is water residence time, $k$ is a unitless parameter representing in-lake P decay, and $x$ is a unitless parameter representing P export via hydrologic flushing. Here, higher values of $k$ are associated with greater integrated P losses from sedimentation and biological uptake resulting in greater P retention. Note that Eq 1 does not include a recycling term. Therefore, our results represent net P retention (as opposed to gross P retention) under a steady state assumption where lakes are at equilibrium with respect to recycling \citep{vollenweiderInputoutputModels1975}. Note that although some forms of the Vollenweider equation use P loading as a predictor variable, it does not appear in Eq 1. The reason for this is two-fold. First, estimates of P loading are more difficult to obtain than estimates of water residence time and our aim was to develop a model than can be widely applied to lakes for which we lack detailed data on P loading. Second, loading based model forms have been shown to be mathematically equivalent to water residence time based model forms \citep{brett2008}.
We used the model described by Eq 1 to compare lake P retention in lakes with different connectivity by fitting the model in two ways. First, we modelled the overall relationship between P retention and water residence time for all lakes in our dataset (global model). Second, we fit hierarchical versions of Eq. 1 where $k$ was modelled separately ($k_{j}$) as a function of a binary sub-population indicator $g_{i}$ denoting membership in one of two lake classes formed on the basis of specific connectivity metrics (or lake depth) and specific spatial extents:

$$
Ri = 1 - \frac{1}{1 + k_{j}t^{x}_{i}}
$$
$$
k_{j} = g_{i}
$$

@@

Our lake connectivity classes were formed by dividing the lake dataset into two classes for each connectivity metric based on the bivariate relationship between each metric and P retention. Prior studies have used a similar binary splitting approach to examine the effect of various exogenous factors on lake P retention \citep{cheng2010, shimodaIntegratingHierarchicalBayes2015}. We determined splitting criteria for each metric from the results of a random forest procedure incorporating conditional inference trees \citep[ctree,]{hothorn2015partykit}. This procedure creates binary splits of the independent variables (i.e. each of the connectivity metrics), which are recursively repeated to find the split that maximizes association with the dependent variable (P retention). The advantage of the ctree technique over the more typical classification-regression tree (CART) technique is that tree growth stopping rules are pre-specified \citep{hothorn2006unbiased}. As a result, some of subjectivity associated with post-hoc tree pruning is avoided.

@@

A secondary purpose of examining the bivariate relationships between connectivity, water residence time, P loading and other factors related to P retention was to determine if our connectivity metrics were related to each other such that they provide similar information. With the exception of stream density and baseflow (r = -0.52, p < 0.05), correlations among connectivity metrics were low and of a similar magnitude as the correlations between connectivity metrics and water residence time (0.10 < r < 0.29). The strongest correlation between any connectivity metric or lake characteristic was upstream lake area and lake watershed area (r = 0.77, p < 0.05).
We observed notably strong correlations between P loading and several lake characteristics not accounted for in Eq 2. These include the correlation between P loading and lake watershed area (r = 0.62, p < 0.05) as well as the correlation between P loading and lake subwatershed area (r = o
 0.76, p < 0.05). In addition, the correlation between P loading and upstream lake area was quite strong (r = 0.52, p < 0.05). Despite strong correlations between P loading, watershed area, and upstream lake area, we did not observe a strong correlation between P loading and P retention (r = 0.15, p > 0.05). Notably, this correlation was much weaker than the correlation between water residence time and P retention (r = 0.44, p > 0.05). Our observation that the correlation between P retention and P loading was not appreciably stronger than correlations between P retention and our connectivity metrics suggests that our estimates of connectivity metric effects on P retention are not confounded by interactions between water residence time and P loading.
Although we did not observe strong correlations among connectivity metrics, we found that lakes with similar connectivity metric values, were spatially clustered (Figure \ref{fig5}). In particular, we found that lakes were concentrated in either the southern or northern portions of our study area depending on their connectivity metric value (Figure \ref{fig5}). This observation is consistent with the findings of \citet{fergusFreshwaterLandscapeLake2017} that lakes in the northern portion of our study area have distinct freshwater connectivity as compared to lakes in the southern portion of our study area.

\begin{figure*}
  \begin{raggedleft}
    \includegraphics[width=0.6\textwidth,keepaspectratio]{../figures/08_maps-1}
  \end{raggedleft}
  \caption{Locations of lakes of with different connectivity metric values at the (A) lake subwatershed and (B) lake watershed extent. Low valued lakes are represented by lighter (green) symbols, and high values lakes are represented by darker (purple) symbols. WS is the lake watershed extent. SWS is the lake subwatershed extent. Connectivity metrics are defined in Figure \ref{fig2}.}\label{fig5}
\end{figure*}

@@

\begin{center}
  \includegraphics[width=0.5\textwidth,keepaspectratio]{../tables/crop/02_model_results_table} \label{table2}
\end{center}

@@

We found that differences in P retention among lake connectivity classes was influenced by specific connectivity metrics including average link length, closest lake distance, and stream density (Table 2, Figure \ref{fig7}). Indeed, these metrics were more strongly associated with P retention than covariates typically used in statistical P retention models (e.g. lake depth, Figure \ref{fig1}). Note that we were able to examine the influence of specific metrics as separate effects apart from water residence time because our model structure treats them as hierarchical coefficients on the P decay parameter $k$. As a result, although water residence time remains the dominant effect on lake P retention, we were able to estimate the specific influence of each metric in a way that is not possible using integrated water residence time and connectivity metrics such as watershed transport capacity \citep{fraterrigoInfluenceLandUse2008}.
Several of the connectivity metrics that were most strongly associated with P retention were weakly correlated with watershed size (Figure \ref{fig4}). The weak nature of these correlations are consistent with the mixed results of prior studies linking watershed size to lake processes. For example, \citet{zimmerLateralVerticalLongitudinal2018} found that carbon export was related to watershed size, but \citet{smithEutrophicationFreshwaterCoastal2003} found that the nitrogen flux was not strongly related to watershed size.
Taken together, our results and the results of prior studies suggest that watershed size and lake depth alone may not always reflect functional differences in potential material transport. One consequence of a correlation between connectivity metrics and watershed size is that metrics derived from watershed size, such as catchment to lake area ratio, are also likely to be associated with connectivity. Catchment to lake area ratio in particular has been previously used as an approximation or proxy of water residence time \citep{rasmussenHumicContentLake1989, sobekPatternsRegulationDissolved2007}.

Although our results differ from those of \citet{soranno2015} who found that the presence of an upstream lake connection was not strongly associated with catchment to lake area ratio, our results suggest that catchment to lake area ratio likely incorporates some connectivity information and caution is needed before using it as a proxy for water residence time.
Another lake characteristic that was strongly related to watershed size was P loading. In particular, positive correlations between P loading and watershed size are consistent with the idea that lakes in larger watersheds receive greater P loading \citep{t-prairieEffectCatchmentSize1986}. A related observation is that the correlation between P loading and upstream lake area was positive. This can be explained by the fact that larger watersheds often have greater numbers of lakes nested within them \citep{zhangEvaluatingEffectsUpstream2012}. Another notable result is that we observed a weaker correlation between P loading and watershed area relative to the correlation between P loading and subwatershed area. One explanation for this result is that land-use cover at the finer lake subwatershed extent has more of an influence on P loading than land-use cover at the broader lake watershed extent \citep{soranno2015}.

@@

We developed a conceptual model that places particulate load fraction in context with other processes affecting P retention (Figure 1). We expected that both land-stream and stream-lake connectivity influences how much particulate P is transported into lakes, which in turn affects their P retention. This expectation is supported by the findings of \citet{cushingTransportDynamicsFine1993} as well as \citet{guy1994} showing that particulate P can be transported beyond the direct drainage from stream-adjacent hillslopes.

>>>>

\begin{figure}
  \begin{center}
    \includegraphics[width=0.4\textwidth, height=3.5in,keepaspectratio]{../figures/01_conceptual_p-cycle_pt2}
  \end{center}
  \caption{Major lake and watershed factors affecting lake P retention. Shaded symbols indicate factors typically considered in P retention models whereas open symbols indicate additional factors specific to the present study. Dashed lines indicate inferred relationships, which cannot be tested with available data, but are otherwise discussed herein.}\label{fig1}
\end{figure}

@@

\begin{figure} \ContinuedFloat
  \begin{center}
    \includegraphics[width=0.5\textwidth,keepaspectratio]{../tables/crop/01_lake_characteristics_table} 
    \captionsetup{labelformat=empty}
    \caption{\textbf{Table 1. } Minimum, median, maximum, and interquartile range of selected characteristics of the study lakes (N = 129).}\label{table1}    
  \end{center}
\end{figure}

@@

In addition to data from the NES and LAGOS-NE datasets, we calculated several connectivity metrics that we expected would be related to lake P retention (Figure \ref{fig2}). Some of these metrics were stream-based with the goal of capturing aspects of the configuration of each lakes’ upstream surface water network (Figure \ref{fig2}C-D). In particular, we chose metrics that would quantitatively approximate network complexity under the assumption that highly complex networks are also low connectivity networks. This assumption is supported by the findings of stream network simulations where increased network complexity leads to increased network resistance and ultimately decreases in network connectivity \citep{rodriguez2001fractal}. In addition to stream-based metrics, we calculated lake-stream-based metrics that we expected would reflect the likelihood of P trapping in upstream lakes prior to arriving at a given focal lake via tributary flow (Figure \ref{fig2}A, 2B). A simple metric that captures this likelihood is the presence (or absence) of an upstream lake (greater than 4 ha) which we define as "lake connection" \citep{fergusFreshwaterLandscapeLake2017}. In addition to lake connection, we calculated related metrics such as total upstream lake area and the network distance to the closest upstream lake.

To examine the importance of spatial extent relative to our connectivity metrics, we calculated each metric at multiple extents (Figure \ref{fig3}). First, we calculated connectivity metrics at the scale of individual lake subwatersheds. We defined a lake subwatershed as the area draining into a particular lake exclusive of any upstream areas that drain into a lake greater than or equal to 10 ha (0.1 $km^{2}$). Next, we calculated connectivity metrics at the scale of entire upstream lake networks (lake watershed extent). We defined a lake watershed as the area draining into any part of the upstream network irrespective of the presence or absence of upstream lakes (Figure \ref{fig3}).

All connectivity metrics were calculated using the high-resolution National Hydrography Dataset (NHD) as a primary input \citep{usgsNationalHydrographyDataset2018}. Average link length was calculated as the total stream length in a given watershed divided by the number of stream reaches after dissolving (removing) any network points that do not occur at a stream junction. Stream density was calculated as the length of all streams in the watershed (minus artificial lines through lakes) expressed in units of meters per hectare. Upstream lake area was calculated as the sum of the lake area in the upstream watershed expressed in square meters. Stream order ratio was defined as the number of headwater (first-order) streams in the upstream watershed of the focal lake divided by the total number of higher order (> 1) streams \citep{labarberaFractalDimensionStream1989}. Closest distance to an upstream lake was defined as the shortest path-distance (rather than the straight-line distance) to a lake upstream from the focal lake.

We calculated all connectivity metrics and lake watershed extents using the `streamnet` and `nhdR` R packages respectively \citep{R-streamnet, R-nhdR}. The algorithms in the `streamnet` package use the `sf` R package \citep{pebesma2018} as well as the `v.net` and `v.stream.order` modules \citep{jasiewiczNewGRASSGIS2011} included in GRASS GIS \citep{GRASS_GIS_software}. All processed connectivity data and code are available at [Zenodo DOI].

@@

We modelled lake total phosphorus retention (hereafter, P retention) using the Vollenweider equation that models P retention as a function of water residence time and a parameter (k) conceptually representing in-lake P decay \citep{chapra2008surface, vollenweiderInputoutputModels1975}. Although there are several variants of this basic equation, we selected a 2-parameter form (equation \ref{eq1}) that has been
shown to have good performance in multiple cross-sectional studies \citep{brett2008, cheng2010}:

\begin{align}
  Ri = 1 - \frac{1}{1 + kt^{x}_{i}} \label{eq1}
\end{align}

where $R_i$ is P retention as a fraction of P inputs, $i$ $\tau$ is water residence time, $k$ is a unitless parameter representing in-lake P decay, and $x$ is a unitless parameter representing P export via hydrologic flushing. Here, higher values of $k$ are associated with greater integrated P losses from sedimentation and biological uptake resulting in greater P retention. Note that Eq \ref{eq1} does not include a recycling term. Therefore, our results represent net P retention (as opposed to gross P retention) under a steady state assumption where lakes are at equilibrium with respect to recycling \citep{vollenweiderInputoutputModels1975}. Note that although some forms of the Vollenweider equation use P loading as a predictor variable, it does not appear in Eq \ref{eq1}. The reason for this is two-fold. First, estimates of P loading are more difficult to obtain than estimates of water residence time and our aim was to develop a model than can be widely applied to lakes for which we lack detailed data on P loading. Second, loading based model forms have been shown to be mathematically equivalent to water residence time based model forms \citep{brett2008}.

We used the model described by Eq 1 to compare lake P retention in lakes with different connectivity by fitting the model in two ways. First, we modelled the overall relationship between P retention and water residence time for all lakes in our dataset (global model). Second, we fit hierarchical versions of Eq. 1 where $k$ was modelled separately ($k_{j}$) as a function of a binary sub-population indicator $g_{i}$ denoting membership in one of two lake classes formed on the basis of specific connectivity metrics (or lake depth) and specific spatial extents:

\begin{align}
  Ri = 1 - \frac{1}{1 + k_{j}t^{x}_{i}} \label{eq2} \\
  k_{j} = g_{i}
\end{align}

@@

Our lake connectivity classes were formed by dividing the lake dataset into two classes for each connectivity metric based on the bivariate relationship between each metric and P retention. Prior studies have used a similar binary splitting approach to examine the effect of various exogenous factors on lake P retention \citep{cheng2010, shimodaIntegratingHierarchicalBayes2015}. We determined splitting criteria for each metric from the results of a random forest procedure incorporating conditional inference trees \citep[][ctree, ]{hothorn2015partykit}. This procedure creates binary splits of the independent variables (i.e. each of the connectivity metrics), which are recursively repeated to find the split that maximizes association with the dependent variable (P retention). The advantage of the ctree technique over the more typical classification-regression tree (CART) technique is that tree growth stopping rules are pre-specified \citep{hothorn2006unbiased}. As a result, some of subjectivity associated with post-hoc tree pruning is avoided.

@@

A secondary purpose of examining the bivariate relationships between connectivity, water residence time, P loading and other factors related to P retention was to determine if our connectivity metrics were related to each other such that they provide similar information. With the exception of stream density and baseflow (r = -0.52, p < 0.05), correlations among connectivity metrics were low and of a similar magnitude as the correlations between connectivity metrics and water residence time (0.10 < r < 0.29). The strongest correlation between any connectivity metric or lake characteristic was upstream lake area and lake watershed area (r = 0.77, p < 0.05).

We observed notably strong correlations between P loading and several lake characteristics not accounted for in Eq 2. These include the correlation between P loading and lake watershed area (r = 0.62, p < 0.05) as well as the correlation between P loading and lake subwatershed area (r = 0.76, p < 0.05). In addition, the correlation between P loading and upstream lake area was quite strong (r = 0.52, p < 0.05). Despite strong correlations between P loading, watershed area, and upstream lake area, we did not observe a strong correlation between P loading and P retention (r = 0.15, p > 0.05). Notably, this correlation was much weaker than the correlation between water residence time and P retention (r = 0.44, p > 0.05). Our observation that the correlation between P retention and P loading was not appreciably stronger than correlations between P retention and our connectivity metrics suggests that our estimates of connectivity metric effects on P retention are not confounded by interactions between water residence time and P loading.

Although we did not observe strong correlations among connectivity metrics, we found that lakes with similar connectivity metric values, were spatially clustered (Figure \ref{fig5}). In particular, we found that lakes were concentrated in either the southern or northern portions of our study area depending on their connectivity metric value (Figure \ref{fig5}). This observation is consistent with the findings of \citet{fergusFreshwaterLandscapeLake2017} that lakes in the northern portion of our study area have distinct freshwater connectivity as compared to lakes in the southern portion of our study area.

\begin{figure*}
  \begin{center}
    \includegraphics[width=0.6\textwidth,keepaspectratio]{../figures/08_maps-1}
  \end{center}
  \caption{Locations of lakes of with different connectivity metric values at the (A) lake subwatershed and (B) lake watershed extent. Low valued lakes are represented by lighter (green) symbols, and high values lakes are represented by darker (purple) symbols. WS is the lake watershed extent. SWS is the lake subwatershed extent. Connectivity metrics are defined in Figure \ref{fig2}.}\label{fig5}
\end{figure*}

@@

\begin{figure*} \ContinuedFloat
  \begin{center}
    \includegraphics[width=0.85\textwidth,keepaspectratio]{../tables/crop/02_model_results_table} 
    \captionsetup{labelformat=empty}
    \caption{\textbf{Table 2. } Connectivity class split values and samples sizes for connectivity metrics and lake depth ranked according to the difference in median $k$ (P decay parameter, $\Delta$k) values. Differences in $\Delta$k that translate to differences in P retention greater than measurement precision are marked with an asterisk. Here, WS is the lake watershed extent whereas SWS is the lake subwatershed extent. Lakes with metric values above or equal to the split value were assigned to a separate connectivity class relative to lakes below the split value. N is sample size. Connectivity metrics are defined in Figure 2.}\label{table2}    
  \end{center}
\end{figure*}

@@

We found that differences in P retention among lake connectivity classes was influenced by specific connectivity metrics including average link length, closest lake distance, and stream density (Table 2, Figure \ref{fig7}). Indeed, these metrics were more strongly associated with P retention than covariates typically used in statistical P retention models (e.g. lake depth, Figure \ref{fig1}). Note that we were able to examine the influence of specific metrics as separate effects apart from water residence time because our model structure treats them as hierarchical coefficients on the P decay parameter $k$. As a result, although water residence time remains the dominant effect on lake P retention, we were able to estimate the specific influence of each metric in a way that is not possible using integrated water residence time and connectivity metrics such as watershed transport capacity \citep{fraterrigoInfluenceLandUse2008}.

Several of the connectivity metrics that were most strongly associated with P retention were weakly correlated with watershed size (Figure \ref{fig4}). The weak nature of these correlations are consistent with the mixed results of prior studies linking watershed size to lake processes. For example, \citet{zimmerLateralVerticalLongitudinal2018} found that carbon export was related to watershed size, but \citet{smithEutrophicationFreshwaterCoastal2003} found that the nitrogen flux was not strongly related to watershed size. Taken together, our results and the results of prior studies suggest that watershed size and lake depth alone may not always reflect functional differences in potential material transport. One consequence of a correlation between connectivity metrics and watershed size is that metrics derived from watershed size, such as catchment to lake area ratio, are also likely to be associated with connectivity. Catchment to lake area ratio in particular has been previously used as an approximation or proxy of water residence time \citep{rasmussenHumicContentLake1989, sobekPatternsRegulationDissolved2007}. Although our results differ from those of \citet{soranno2015} who found that the presence of an upstream lake connection was not strongly associated with catchment to lake area ratio, our results suggest that catchment to lake area ratio likely incorporates some connectivity information and caution is needed before using it as a proxy for water residence time.

Another lake characteristic that was strongly related to watershed size was P loading. In particular, positive correlations between P loading and watershed size are consistent with the idea that lakes in larger watersheds receive greater P loading \citep{t-prairieEffectCatchmentSize1986}. A related observation is that the correlation between P loading and upstream lake area was positive. This can be explained by the fact that larger watersheds often have greater numbers of lakes nested within them \citep{zhangEvaluatingEffectsUpstream2012}. Another notable result is that we observed a weaker correlation between P loading and watershed area relative to the correlation between P loading and subwatershed area. One explanation for this result is that land-use cover at the finer lake subwatershed extent has more of an influence on P loading than land-use cover at the broader lake watershed extent \citep{soranno2015}.

@@

We developed a conceptual model that places particulate load fraction in context with other processes affecting P retention (Figure \ref{fig1}). We expected that both land-stream and stream-lake connectivity influences how much particulate P is transported into lakes, which in turn affects their P retention. This expectation is supported by the findings of \citet{cushingTransportDynamicsFine1993} as well as \citet{guy1994} showing that particulate P can be transported beyond the direct drainage from stream-adjacent hillslopes.
